---
title: "Reproducible Research, Project 1"
author: "Poul Kristensen"
date: "April 1, 2018"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Load packages and data

```{r}
library(ggplot2)
library(plyr)

steps <- read.csv('activity.csv')
summary(steps)
```

Get the total number of days and steps in sample, and calculate the mean number of steps per day:

```{r}
Ndays <- length(unique(steps$date))
Nsteps <- sum(steps$steps[!is.na(steps$steps)])
Meansteps <- Nsteps / Ndays
print(Ndays)
print(Nsteps)
print(Meansteps)

```

Use the ddply function to aggregate steps by day:

```{r}
stepsbyday <- ddply(steps, .(date), summarize, countbyday = sum(steps, na.rm = TRUE) )

```

Create a histogram of steps per day:

```{r}
qplot(stepsbyday$countbyday, geom="histogram", xlab = "Steps per day") 

```

Calculate and report the mean and median of steps per day:

```{r}
summary(stepsbyday$countbyday)

```

Calculate the average number of steps by 5-minute interval within the day, across all days, then plot the time series:

```{r}
byinterval <- ddply(steps[!is.na(steps$steps),], .(interval), summarize, meansteps = mean(steps))
plot(unique(steps$interval), byinterval$meansteps, type="l", xlab="Time interval",
  	ylab="Mean number of steps within interval")  

```

Find and print the 5-minute interval with the highest average number of steps:

```{r}
i <- which(max(byinterval$meansteps) == byinterval$meansteps)
print(byinterval$interval[i])
```

Calculate and print the total number of missing values:

```{r}
Nmissing <- sum(is.na(steps$steps))
print(Nmissing)
```


Replace missing values for steps by the average for that 5-minute interval across days:

```{r}
itoimpute <- is.na(steps$steps)          #identify missing values
stepsNew <- steps
stepsNew$stepsImputed <- steps$steps   

for (i in 1:length(itoimpute)) { 
  if (itoimpute[i] == TRUE) {
    stepsNew$stepsImputed[i] <- mean(steps$steps[!is.na(steps$steps) & steps$interval == steps$interval[i]], rm.na = TRUE)
    }
}
```

Again, calculate the total number of steps per day, using the ddply function, but now in the new dataset with imputed values:

```{r}
stepsbydayNew <- ddply(stepsNew, .(date), summarize, countbyday = sum(stepsImputed, na.rm = TRUE) )
```

Display histogram of steps per day:

```{r}
qplot(stepsbydayNew$countbyday, geom="histogram", xlab = "Steps per day") 
```

Calculate and display the mean and median of steps per day in the new dataset:

```{r}
summary(stepsbydayNew$countbyday)
```

When imputing missing data, the mean and median number of steps per day are not significantly  different, although the mean is a bit higher. The small difference is explained by the fact that all of the missing values are for the first 8 days of the time period, thus evenly distributed across times within the day. Missing values during the first days of the sample would only affect the estimated pattern of steps through the day if there had been a significant trend in overall step activity through the days of the sample.

Convert date to proper date format and create "weekday"/"weekend" factor variable:

```{r}
steps$date2 <- as.Date(steps$date, format = "%Y-%m-%d")
steps$dayofweek <- weekdays(steps$date2)
steps$WEornot[steps$dayofweek == "Saturday" | steps$dayofweek == "Sunday"] = "weekend"
steps$WEornot[is.na(steps$WEornot)] = "weekday"
steps$WEornot <- as.factor(steps$WEornot)
```

Calculate the average number of steps by time interval for weekdays and weekends, then create panel plot for weekdays and weekends:

```{r}
byinterval <- ddply(steps[!is.na(steps$steps),], .(interval, WEornot), summarize, meansteps = mean(steps))

par(mfrow = c(2,1))
plot(byinterval$interval[byinterval$WEornot == "weekend"], byinterval$meansteps[byinterval$WEornot == "weekend"], type="l", main = "weekend", xlab="Time interval", 	ylab="Mean number of steps within interval")  
plot(byinterval$interval[byinterval$WEornot == "weekday"], byinterval$meansteps[byinterval$WEornot == "weekday"], type="l", main = "weekday", xlab="Time interval", 	ylab="Mean number of steps within interval")  
```




